Estimating SARS-CoV-2 transmission in England from randomly sampled prevalence surveys

### 2024-02-09 transmission.tex
finish split into appendix

\Textcite{osogamiClosed} provide a method to construct the compartments, implemented in the mapfit R package~\autocite{mapfit} implements the expressions to calculate the EC distribution parameters given a distribution's first three moments.
Applying the package to the posterior mean of the distribution estimated in \cref{E-imperf-test} gives the model in \cref{SEIR:fig:full-model} with the parameter estimates in \cref{SEIR:table:ec-params} (see \cref{transmission:sec:phase-type} for details).
\begin{table}
    \centering
    \begin{tabular}{c c c c c}
        $l$ & $\omega_Y$ & $\omega_{X1}$ & $\omega_{X2}$ & $\pi$ \\
        3 & 0.0820 & 0.126 & 0.0223 & 0.00973  \\
    \end{tabular}
    \caption{Parameter estimates for the EC distribution approximating the posterior mean of the duration distribution estimated in \cref{E-imperf-test}.}
    \label{SEIR:table:ec-params}
\end{table}

The mapfit R package~\autocite{mapfit} implements the expressions to calculate the EC distribution parameters given a distribution's first three moments.
Applying the package to the posterior mean of the distribution estimated in \cref{E-imperf-test} gives the parameter estimates in \cref{SEIR:table:ec-params}.
\begin{table}
    \centering
    \begin{tabular}{c c c c c}
        $l$ & $\omega_Y$ & $\omega_{X1}$ & $\omega_{X2}$ & $\pi$ \\
        3 & 0.0820 & 0.126 & 0.0223 & 0.00973  \\
    \end{tabular}
    \caption{Parameter estimates for the EC distribution approximating the posterior mean of the duration distribution estimated in \cref{E-imperf-test}.}
    \label{SEIR:table:ec-params}
\end{table}
% end table of ec params

I link the modelled proportion of individuals that are RT-PCR positive to the data using a beta-binomial likelihood (as discussed in \cref{transmission:sec:data}).
If on each day $t$, each strata $i$ is observed to have $y_{at}$ positives out of $n_{at}$ tests, then the likelihood is:
\begin{align}
    \prod_{a,t} \BB (y_{at} \mid n_{at}, \bar{p}_{at}, \rho)
    \label{SEIR:eq:likelihood}
\end{align}
where: $\bar{p}_{at}$ is the mean proportion of individuals in strata $a$ that are RT-PCR positive on day $t$; and $\rho$ is a nuisance parameter controlling the overdispersion of the beta-binomial distribution.
Denoting the proportion of stratum $a$ in state $p_k$ as $p_{ka}(\tau)$, then $\bar{p}_{at} = \int_{t}^{t+1} \sum_{k=1}^l p_{ka}(\tau) d\tau$.

@@

    \path  (S) edge node {$\lambda$} (P0)
          (P0) edge node {$\omega_0$} (P1)
          (P1) edge node {$\omega_Y$} (P2)
          (P2) edge node {$\pi \omega_{X1}$} (P3)
               edge [out=-45,in=-135] node[below] {$(1 - \pi) \omega_{X1}$} (P4)
          (P3) edge node {$\omega_{X2}$} (P4);

@@

A diagram of the model for a single stratum is shown in figure \cref{SEIR:fig:full-model}.
The full system of differential equations for each region is below.
\begin{align}
    \label{SEIR:eq:fullODEs}
    \vec{\lambda}(t) &= \beta(t) (\matr{\kappa}(t) \circ \matr{q}) ((\vec{i_1}(t) + \vec{i_2}(t)) \circ \vec{N}) \\
    \frac{d\vec{s}(t)}{dt} &= -\vec{\lambda}(t) \circ \vec{s}(t) \\
    \frac{d\vec{e_1}(t)}{dt} &= \vec{\lambda}(t) \circ \vec{s}(t) - 2\sigma \vec{e_1}(t) \\
    \frac{d\vec{e_2}(t)}{dt} &= 2\sigma \vec{e_1}(t) - 2\sigma \vec{e_2}(t) \\
    \frac{d\vec{i_1}(t)}{dt} &= 2\sigma \vec{e_2}(t) - 2\gamma \vec{i_1}(t) \\
    \frac{d\vec{i_2}(t)}{dt} &= 2\gamma \vec{i_1}(t) - 2\gamma \vec{i_2}(t) \\
    \frac{d\vec{r}(t)}{dt} &= 2\gamma \vec{i_2}(t) \\
    \frac{d\vec{p_0}(t)}{dt} &= \vec{\lambda} \circ \vec{s}(t) - \omega_0 \vec{p_0}(t) \\
    \frac{d\vec{p_1}(t)}{dt} &= \omega_0 \vec{p_0}(t) - \omega_Y \vec{p_1}(t) \\
    \frac{d\vec{p_2}(t)}{dt} &= \omega_Y \vec{p_1}(t) - \omega_{X1} \vec{p_2}(t) \\
    \frac{d\vec{p_3}(t)}{dt} &= \pi \omega_{X1} \vec{p_2}(t) - \omega_{X2} \vec{p_3}(t)
\end{align}
where $\circ$ indicates element-wise multiplication.

@@

From the system of ODEs, the knowledge that all the compartments are growing exponentially at rate $\psi$, and the parameters $\vec{\pi}$ (where $\pi_a$ is the proportion of strata $i$ without immunity at time 0, hence, $\vec{r} = 1 - \vec{\pi}$) one can derive the following for the initial state of the compartments:
\begin{align}
    \vec{i_1} &= \vec{i_0} \left(1 + \frac{2\gamma}{\psi + 2\gamma} \right)^{-1} \\
    \vec{i_2} &= \vec{i_0} - \vec{i_1} \\
    \vec{e_2} &= \vec{i_1} \frac{\psi + 2\gamma}{2\sigma} \\
    \vec{e_1} &= \vec{e_2} \frac{\psi + 2\sigma}{2\sigma} \\
    \vec{s} &= \vec{\pi} - \vec{i_0} - \vec{e_1} - \vec{e_2} \\
    \vec{p_0} &= \vec{e_1} \frac{\psi + 2\sigma}{\psi + \omega_0} \\
    \vec{p_1} &= \vec{p_0} \frac{\omega_0}{\psi + \omega_Y} \\
    \vec{p_2} &= \vec{p_1} \frac{\omega_Y}{\psi + \omega_{X1}} \\
    \vec{p_3} &= \vec{p_2} \frac{\pi \omega_{X1}}{\psi + \omega_{X2}}.
\end{align}

>>>>

\Textcite{osogamiClosed} provide a method to construct the compartments, implemented in the mapfit R package~\autocite{mapfit}. 
The resulting compartmental approximation matches the original distribution up to the first three moments.
Applying the package to the posterior mean of the distribution estimated in \cref{E-imperf-test} gives the model in \cref{SEIR:fig:full-model} with the parameter estimates in \cref{SEIR:table:ec-params} (see \cref{transmission:sec:phase-type} for details).
\begin{table}
    \centering
    \begin{tabular}{c c c c c}
        $l$ & $\omega_1$ & $\omega_{2}$ & $\omega_{3}$ & $\pi$ \\
        3 & 0.0820 & 0.126 & 0.0223 & 0.00973  \\
    \end{tabular}
    \caption{Parameter estimates for the EC distribution approximating the posterior mean of the duration distribution estimated in \cref{E-imperf-test}.}
    \label{SEIR:table:ec-params}
\end{table}
% end table of ec params

I link the modelled proportion of individuals that are RT-PCR positive to the data using a beta-binomial likelihood (as discussed in \cref{E-biology-data:sec:positivity}).
If on each day $t$, each strata $i$ is observed to have $y_{at}$ positives out of $n_{at}$ tests, then the likelihood is:
\begin{align}
    \prod_{a,t} \BB (y_{at} \mid n_{at}, \bar{p}_{at}, \rho)
    \label{SEIR:eq:likelihood}
\end{align}
where: $\bar{p}_{at}$ is the mean proportion of individuals in strata $a$ that are RT-PCR positive on day $t$; and $\rho$ is a nuisance parameter controlling the overdispersion of the beta-binomial distribution.
Denoting the proportion of stratum $a$ in state $p_k$ as $p_{ka}(\tau)$, then $\bar{p}_{at} = \int_{t}^{t+1} \sum_{k=1}^l p_{ka}(\tau) d\tau$.

@@

    \path  (S) edge node {$\lambda$} (P0)
          (P0) edge node {$\omega_0$} (P1)
          (P1) edge node {$\omega_1$} (P2)
          (P2) edge node {$\pi \omega_{2}$} (P3)
               edge [out=-45,in=-135] node[below] {$(1 - \pi) \omega_{2}$} (P4)
          (P3) edge node {$\omega_{3}$} (P4);

@@

A diagram of the model for a single stratum is shown in figure \cref{SEIR:fig:full-model}.
The full system of differential equations for each region is below.
\begin{align}
    \label{SEIR:eq:fullODEs}
    \vec{\lambda}(t) &= \beta(t) (\matr{\kappa}(t) \circ \matr{q}) ((\vec{i_1}(t) + \vec{i_2}(t)) \circ \vec{N}) \\
    \frac{d\vec{s}(t)}{dt} &= -\vec{\lambda}(t) \circ \vec{s}(t) \\
    \frac{d\vec{e_1}(t)}{dt} &= \vec{\lambda}(t) \circ \vec{s}(t) - 2\sigma \vec{e_1}(t) \\
    \frac{d\vec{e_2}(t)}{dt} &= 2\sigma \vec{e_1}(t) - 2\sigma \vec{e_2}(t) \\
    \frac{d\vec{i_1}(t)}{dt} &= 2\sigma \vec{e_2}(t) - 2\gamma \vec{i_1}(t) \\
    \frac{d\vec{i_2}(t)}{dt} &= 2\gamma \vec{i_1}(t) - 2\gamma \vec{i_2}(t) \\
    \frac{d\vec{r}(t)}{dt} &= 2\gamma \vec{i_2}(t) \\
    \frac{d\vec{p_0}(t)}{dt} &= \vec{\lambda} \circ \vec{s}(t) - \omega_0 \vec{p_0}(t) \\
    \frac{d\vec{p_1}(t)}{dt} &= \omega_0 \vec{p_0}(t) - \omega_1 \vec{p_1}(t) \\
    \frac{d\vec{p_2}(t)}{dt} &= \omega_1 \vec{p_1}(t) - \omega_{2} \vec{p_2}(t) \\
    \frac{d\vec{p_3}(t)}{dt} &= \pi \omega_{2} \vec{p_2}(t) - \omega_{3} \vec{p_3}(t)
\end{align}
where $\circ$ indicates element-wise multiplication.

@@

From the system of ODEs, the knowledge that all the compartments are growing exponentially at rate $\psi$, and the parameters $\vec{\pi}$ (where $\pi_a$ is the proportion of strata $i$ without immunity at time 0, hence, $\vec{r} = 1 - \vec{\pi}$) one can derive the following for the initial state of the compartments:
\begin{align}
    \vec{i_1} &= \vec{i_0} \left(1 + \frac{2\gamma}{\psi + 2\gamma} \right)^{-1} \\
    \vec{i_2} &= \vec{i_0} - \vec{i_1} \\
    \vec{e_2} &= \vec{i_1} \frac{\psi + 2\gamma}{2\sigma} \\
    \vec{e_1} &= \vec{e_2} \frac{\psi + 2\sigma}{2\sigma} \\
    \vec{s} &= \vec{\pi} - \vec{i_0} - \vec{e_1} - \vec{e_2} \\
    \vec{p_0} &= \vec{e_1} \frac{\psi + 2\sigma}{\psi + \omega_0} \\
    \vec{p_1} &= \vec{p_0} \frac{\omega_0}{\psi + \omega_1} \\
    \vec{p_2} &= \vec{p_1} \frac{\omega_1}{\psi + \omega_{2}} \\
    \vec{p_3} &= \vec{p_2} \frac{\pi \omega_{2}}{\psi + \omega_{3}}.
\end{align}
